---
title: "GPS Track Demo"
author: "Brian High"
date: "11/12/2015"
output:
  html_document:
    fig_caption: no
    fig_width: 5
    keep_md: yes
---

This is one way to plot a track from a GPS file (GPX, TCX, etc.) file using 
R and `ggmap`. 

Based on: 
[Mapping GPS Tracks in R](http://www.r-bloggers.com/mapping-gps-tracks-in-r/) 
by Shawn Mollie Taylor. Here is the 
[Gist](https://gist.github.com/mollietaylor/4210660).

## Set Options

First we will set a few `knitr` options to control document rendering.

```{r, echo=TRUE, results='asis', warning=FALSE, cache=FALSE, message=TRUE}
suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE)
```

## Install Packages

Then we will install `ggmap` if it is not already installed and load the 
package.

```{r}
if (!require(ggmap)) install.packages("ggmap", repos="http://cran.fhcrc.org")
library(ggmap)
```

## Convert GPS File to CSV

Now, we convert our GPS file to a CSV for importing into R using 
[GPSBabel](http://www.gpsbabel.org/). 

This particular GPS track was created from a smartphone using the MapMyWalk app 
and downloaded as a TCX file from the [MapMyWalk](www.mapmywalk.com) website.

The following command is run from the "shell" prompt of the computer's 
operating system, such as Bash, DOS, or PowerShell.

```{r, engine="bash", eval=FALSE}
gpsbabel -t -i gtrnctr -f mywalk.tcx -o unicsv -F mywalk.csv
```

## Import the CSV

We can import this CSV file into R with `read.csv`.

```{r}
gps <- read.csv('mywalk.csv', header = TRUE)
```

## Get the Base Map

### Calculate Route Location and Size

To get the right base map, we need to perform some calculations. We want to 
know the center coordinates and the width and height of the route.

```{r}
centerLon <- median(gps$Longitude)
centerLat <- median(gps$Latitude)
routeWidth <- 2 * abs(max(gps$Longitude) - min(gps$Longitude))
routeHeight <- 2 * abs(max(gps$Latitude) - min(gps$Latitude))
```

### Google Map

We can get our base map from [Google](https://www.google.com/maps). The zoom 
level was determined by trial-and-error, knowing the range should be between
3 (continent) and 21 (building) with 10 suggested for a "city" zoom level.

```{r}
zoomlevel <- 15
googleMapImageData <- get_googlemap(
    center = c(lon = centerLon, lat = centerLat), 
    zoom = zoomlevel, maptype = c("terrain")
    )
```

### OpenStreetMap

Or we can use a base map from [OpenStreetMap](https://www.openstreetmap.org/). 
The value we chose for scale approximates the zoom chosen for the Google Map
and was based on a [formula](http://gis.stackexchange.com/questions/7430) 
found online for converting zoom level to map scale. 

```{r}
openStreetMapImageData <- get_openstreetmap(
    bbox = c(lowerleftlon=centerLon-routeWidth/2, 
             lowerleftlat=centerLat-routeHeight/2, 
             upperrightlon=centerLon+routeWidth/2, 
             upperrightlat=centerLat+routeHeight/2),
    scale = floor(591657550.5 / 2^(zoomlevel - 1) / 2)
    )
```

## Plot the Track on the Map

### Google Map

Here is the GPS track plotted over a Google base map.

```{r}
ggmap(googleMapImageData, extent = "device") + 
    geom_point(aes(x = Longitude, y = Latitude), data = gps,
               colour = "red", size = 1, pch = 20)
```

### OpenStreetMap

And here is the same GPS track plotted on an OpenStreetMap.

```{r}
ggmap(openStreetMapImageData, extent = "device") + 
    geom_point(aes(x = Longitude, y = Latitude), data = gps,
               colour = "red", size = 1, pch = 20)
```